\chapter*{Acknowledgements}
%\chapter*{Agradecimentos}

Thank you,

\flushleft{My family: to my parents for fostering me to be curious and knowledgeable, and helping me explore the world over the years; to my godparents for supporting me and being the stern voice when I am too bohemian; to my brother, for making me angry, laugh, and cry with joy.}

\flushleft{My friends from Matosinhos: to Luís, Francisco, Pedro, Diogo, Gonçalo, Filipa, Tiago, Tozé, Matilde, and Martins, for sharing so many years and important milestones. To Eduardo and Leonor for putting up with me in our adventures and staying close to me no matter what.}

\flushleft{My friends from university: to Bernas, Vítor, Luís, Miguel, Cajó, Pedro, Tito, for being the best mates I could ask, right from the start. To Manel, the one that got away. To André, for being my role model and inspiration to always do better in our field.}

\flushleft{To João Macedo, for being my brother in arms. For the times we spent as teens, for the trips we made, for the miles we walked together, for the calls, for the nights out bar crawling and the nights in pulling all-nighters coding, for Berlin and Valencia, for all else we shared. You started as a drop-in from Computer Engineering, and now look where you are. I've seen you grow so much, and grown up so much myself, and I cannot shake how much of a privilege it is to have you in my life.}

\flushleft{To professor João Bispo for supervising my thesis work. For being the approachable professor in Compilers and Software Language Engineering, for being enthusiastic in the work I was doing and pushing me to consider the broader context and consequences of it. For being my editor, correcting and nitpicking and making me a better author.}

\flushleft{To David, Teté, Pedro Nuno, Carolina Losada, Bernardo Moreira, Luís Sousa, and so many others.}

\flushleft{Yours sincerely,}
\flushleft{João}
